---
title: "A Brief Intro to `inferference`"
author: 
- name : "Bradley Saul"
  affiliation: "University of North Carolina"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inferference_intro}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Use of `inferference` 

This is a brief description of how to use the `inferference` package.

```{r, echo = TRUE}
# load the package
library(inferference)
```

I have included a simple dataset generated from `rinterferenceSim` package: 3000 units, 250 groups, 2 covariates, 21 allocations (0 to 1 by 0.05), using the parameters in the code below. See Perez 2014 for details on how the simulations were structured.


```{r, echo=TRUE, eval=FALSE}
alphaz = seq(0, 1, by=.05) # needed to compute truth
theta.base <- c(.5, -0.788, -2.953, -0.098, -0.145, 0.351)
theta.sim <- c(0.2727, -0.0387, .2719, 1.0859)
sample_sim <- sim_interference(n = 3000, N = 250, nsims = 1, 
                               base.parameters = theta.base, 
                               parameters = theta.sim,
                               alphas = alphaz)
interference_sample <- sample_sim$sims[[1]]
#save(sample_data, file = "data/sample_data.rda")
```

Here's what the data looks like:
```{r}
head(vaccinesim)
```

y is the outcome. X1 and X2 are covariates. A is the randomized treatment indicator. B is an indicator of participation in the trial. Group is obvious.

# `interference` function

The primary function in `inferference` is `interference`. Currently, it only supports IPW estimation. `interference` has several options:

* `propensity_integrand`: a character string that names the function that defines the group-level propensity: $P(\mathbf{A}_i | \mathbf{X}_i)$. The default function is `logit_integrand`. This function handles models with and without random effects. Without random effects, `logit_integrand` computes:
\[
\prod_{j = 1}^n_i h_{ij}^{A_{ij}}(1 - h_{ij})^{1 - A_{ij}}
\]
where $h_{ij} = logit^{-1}(\mathbf{X}\hat{\theta})$; $\hat{\theta}$ are the estimated parameters from logistic regression using `glm`. In the case of a random effect, `logit_integrand` computes:
\[
\prod_{j = 1}^n_i h_{ij}(b_i)^{A_{ij}}(1 - h_{ij}(b_i))^{1 - A_{ij}}f_b(b_i)
\]
where $h_{ij} = logit^{-1}(\mathbf{X}\hat{\theta} + b)$; $\hat{\boldsymbol{\theta}}_f$ are the estimated fixed effect parameters from random effects logistic regression using `glmer`; $f_b(\cdot)$ is a $N(0, \hat{\theta}_r)$ density; $\hat{\theta}_r$ is the estimated random effect variance. For both cases `logit_integrand` is passed through the `integrate` function. Without a random effect, however, this has no effect since `integrate(function(b) dnorm(b))` equals 1.
* formula: this has a minimum of 4 parts in the following structure:
  * `outcome | exposure ~ propensity covariates | group`. 
If a random effect for the group is included use the following: 
  * `outcome | exposure ~ propensity covariates + (1|group) | group`.
If the study design includes a 'participation' variable like in Perez-Heydrich 2014 use the following:
  * `outcome | exposure | participation ~ propensity covariates + (1|group) | group`.
* allocations: a vector of treatment allocation strategies. At least 2 allocations must defined.
* data: the data.frame with the outcome, exposure, covariates, and group variables
* model_method: one of `oracle`, `glm`, or `glmer`.
* model_options: options passed to the `model_method` as a list. For example, for `glm` or `glmer` set this equal to `list(family = binomial(link = 'logit'))`. To use the `oracle` method, use `list(fixed.effects = c(???, ???, etc), random.effects = NULL)` or `list(fixed.effects = c(???, ???, etc), random.effects = ???)`
* causal_estimation_options: list of options passed to `ipw_interference`. `list(variance_estimation = 'simple')` or `list(variance_estimation = 'robust')` are the option settings of most interest. 'robust' is the default.
'robust' uses M-estimation theory to compute the variance, whereas, 'simple' computes $\frac{1}{N^2}\sum_{i= 1}^N (\hat{CE}_i - \hat{CE})^2$, where $\hat{CE}$ is point estimate for the causal effect of interest while $\hat{CE}_i$ is the point estimate for group $i$.
* conf.level: the confidence level. defaults to .95.
* rescale.factor: a scaling factor for the causal effects. Defaults to 1.
* ... other options passed to `grad()`, `integrate()`, or the propensity_integrand function. `method = simple` will vastly speed up the `grad()` function. 

# Examples

Here's a few samples of `interference()` based on the data provided. 

## Replication of Perez-Heydrich 2014

The first example follows Perez-Heydrich 2014 and includes a random effect and participation variable. Conditional on participation, subjects were randomized to treatment with probability 2/3, which is the `randomization` argument. This is an argument in `logit_integrand()`. This argument defaults to $1$ and so can be left out if unnecessary (see later examples). 

```{r, echo=TRUE}
sample1 <- interference(
  allocations = c(.3, .45,  .6), # a minimum of 2 is required
  formula = y | A | B ~ X1 + X2 + (1|group) | group, 
  data = vaccinesim, # name of the data frame
  randomization = 2/3,
  method = 'simple' # speeds up grad()
  )
```

Take a quick look using the (unpolished) `print` function:
```{r, echo = TRUE}
print(sample1)
```

Other pieces that might be interest are the group-level weights:
\[
\frac{\prod_{j =1}^{n_i} \alpha^{A_{iij}}(1 - \alpha)^{1 - A_{ij}}}{P(\mathbf{A}_i | \mathbf{X}_i)}
\]
which can accessed:
```{r, echo = TRUE}
# the matrix of weights
head(sample1$weights)
```

The weight 'derivatives':
\[
\frac{\prod_{j =1}^{n_i} \alpha^{A_{iij}}(1 - \alpha)^{1 - A_{ij}}}{\frac{d}{d\theta}P(\mathbf{A}_i | \mathbf{X}_i)}
\]
which can accessed:
```{r, echo = TRUE}
head(sample1$weightd[ , , 1]) # For the 1st allocation 
```

The 'scores' are but the all the last element in $\psi(\mathbf{O}, \boldsymbol{\theta})$, eqn (3) in the web appendix, which can accessed:
```{r, echo = TRUE}
head(sample1$scores) 
```

Finally, the most important piece, the effect estimates are in a data.frame:
```{r, echo = TRUE}
sample1$estimates 
```

Helper functions for finding effects:
```{r, echo=TRUE}
direct_effect(sample1, .3) #DE(.3)/ unfortunately de() is a function in utils.
ie(sample1, .3, .6) #IE(.3, .6) 
indirect_effect(sample1, .3, .6) # same
te(sample1, .3, .6) #TE(.3, .6)
oe(sample1, .3, .6) #OE(.3, .6)

# You can use the same functions to get all the effects. This will aid in 
# plotting (see below.)

# All direct effect of constrast Y(0) - Y(1)
direct_effect(sample1) 

# to get Y(1) - Y(0) use trt.lvl1 argument
direct_effect(sample1, trt.lvl1 = 1) 

# Other functions work similarly

# all indirect effects where treatment = 0 compare to an allocation of .3
ie(sample1, .3) 
```
See `?direct_effect`, `?ie`, `?te`, or `?oe` for options.


## Without random effects
```{r, echo=TRUE}
sample2 <- interference(
  allocations = c(.3, .6), # a minimum of 2 is required
  formula = y | A | B ~ X1 + X2 | group,
  data = vaccinesim, # name of the data frame
  model_method = 'glm',
  model_options = list(family = binomial),
  randomization = 2/3,
  method = 'simple' # speeds up grad()
  )
print(sample2)
```

## Using oracle parameters w/o the random effects
```{r, echo=TRUE}
sample3 <- interference(
  allocations = c(.3, .6), # a minimum of 2 is required
  formula = y | A | B ~ X1 + X2 | group,
  data = vaccinesim, # name of the data frame
  model_method = 'oracle',
  model_options = list(fixed.effects = c(0.2727, -0.0387, .2719),
                       random.effects = NULL),
  causal_estimation_options = list(variance_estimation = 'simple'),
  randomization = 2/3,
  method = 'simple' # speeds up grad()
  )
print(sample3)
```

## Without the participation variable

Suppose we don't have the participation variable and estimate the treatment probability directly, just leave out the `randomization` argument and modify the formula:
```{r, echo=TRUE}
sample4 <- interference(
  allocations = c(.3, .6), # a minimum of 2 is required
  formula = y | A ~ X1 + X2 | group,
  data = vaccinesim, # name of the data frame
  model_method = 'glm',
  causal_estimation_options = list(variance_estimation = 'robust'),
  method = 'simple' # speeds up grad()
  )
print(sample4)
```

## Defining one's own integrand

Let's say you want to fix the group level propensity to be .5. I don't know why you'd want to do this, but this illustrates how to write your own function. The function required three arguments: `b`, `x`, and `pos`. `b` must be the first argument. `x` and `pos` are necessary for the `grad()`, and  I will describe how to use these later.

```{r, echo = TRUE}
myFUN <- function(b, x, pos){
  return(.5 * dnorm(b))
}

integrate(myFUN, l = -Inf, u = Inf) # returns 0.5

sample5 <- interference(
  propensity_integrand = 'myFUN',
  allocations = c(.3, .6), # a minimum of 2 is required
  formula = y | A ~ X1 + X2 | group,
  data = vaccinesim, # name of the data frame
  model_method = 'glm',
  causal_estimation_options = list(variance_estimation = 'simple'),
  # won't work for 'robust' unless addition arguments are defined
  method = 'simple' # speeds up grad()
  )
print(sample5)
```

## A sample plot

Here's a simple plot of direct effects. Run the `interference` function with a larger number of allocations (this will take longer!):

```{r, echo = TRUE}
sample6 <- interference(
  allocations = seq(.3, .6, by = .01), 
  formula = y | A ~ X1 + X2 | group,
  data = vaccinesim, # name of the data frame
  model_method = 'glm',
  causal_estimation_options = list(variance_estimation = 'robust'),
  method = 'simple' # speeds up grad()
  )

# Look at the direct effects
deff <- direct_effect(sample6)

# Plot the point estimates
plot(deff$alpha1, deff$estimate, type = 'l')
```
